---
title: "tidy_data_class3"
output: html_document
date: "2026-09-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(tidyverse)
library(readxl)
library(haven)
```
# loading datasets
```{r}
pulse_df =
  haven::read_sas("data_file/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()
```


# tidying data 

take column names and put them into a new variable, take values and put them into a new variable

```{r}
pulse_tidy_df = 
  pulse_df |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", # where should you send the column names
    names_prefix = "bdi_score_",
    values_to = "bdi_score" # where should you send the values, to a column
  ) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  )

```

# practicing with litters data

```{r}
pulse_df =
  read_csv("data_file/FAS_litters.csv") |> 
  janitor::clean_names() |> 
  select(litter_number, gd0_weight, gd18_weight) |> 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd",
    values_to = "weight"
  ) |> 
  mutate(
    gd_number = recode_values(
      gd,
      "gd0_weight" ~ 0,
      "gd18_weight" ~ 18,
    )
  )
```

#untidying data

```{r}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
  )
analysis_result
```


pull values and put them into new column

Check this code:

analysis_result |> 
  pivot_wider(
    names_from = time,
    values_from = mean_outcome
  ) |> 
  knitr::kable()
  
```{r}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
  
```

#new example

```{r}
fellowship_df = 
  readxl::read_excel("data_file/LotR_Words.xlsx", range = "B3:D6") |> 
  mutate(movie = "fellowship")

# copy and pasting the code and changing df name and cell ranges from same excel file

two_towers_df = 
  readxl::read_excel("data_file/LotR_Words.xlsx", range = "F3:H6") |> 
  mutate(movie = "two towers")

return_df = 
  readxl::read_excel("data_file/LotR_Words.xlsx", range = "J3:L6") |> 
  mutate(movie = "return")

#once you have copy and pasted your code more than twice, you should write a function to do it and make it easier (we will do this later)

```

Next, let's put it all together and tidy
```{r}
lotr_df = 
  bind_rows(fellowship_df, two_towers_df, return_df) |>
  janitor::clean_names() |> 
  relocate(movie) |> 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"
  )
```

## join FAS datasets

```{r}
pups_df = 
  read_csv(
    "data_file/FAS_pups.csv",
    skip = 3,
    na = c(".", "", "NA")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = recode_values( # recoding variables, 1 is male and 2 is female (double check case_match vs recode_variables)
      sex,
      1 ~ "male",
      2 ~ "female"
    )
  )

litters_df = 
  read_csv(
    "data_file/FAS_litters.csv",
    na = c(".", "", "NA")) |> 
  janitor::clean_names() |> 
  relocate(litter_number) |>  # want litter number to appear first
  separate(group, into = c("dose", "day_of_tx"), 3) |> # split at the third character
  mutate(
    dose = str_to_lower(dose),
    day_of_tx = as.numeric(day_of_tx),
    gd_weight_gain = gd18_weight - gd0_weight
  )


fas_df =
  left_join(pups_df, litters_df, by = "litter_number") # tell R what columns are the id variables

# multiple pups have the same litter number, it will just duplicate those rows for you

```





